A recipe for extremely reproducible enrichment analysis

### 2023-03-09 manuscript.Rmd
mswork

---
title: "A recipe for extremely reproducible enrichment analysis"
Enrichment analysis with gold standard reproducibility"
author: "Mark Ziemann and Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

@@

The lack of reproducibility is particularly acute for enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script based methods 
[5](Wijesooriya et al, 2022).
Web-based tools are a particular concern, as it is known that algorithms and functional annotation
sets undergo regular updates, but the versions are not regularly reported nor captured, 
making later replication difficult if not impossible [5](Wijesooriya et al, 2022).

A lack of reproducibility does not mean a study is invalid, however it cannot be ruled out that
mistakes have been made in the enrichment analysis and that the results are biased or false.
As omics data are becoming more routine in the drug development process and are used to
inform clinical practice, reproducibility should become a minimum requirement for 
bioinformatics analysis of omics data.

Although it doesn't need to be this way due to the precise nature of modern computers and software.
In order for bioinformatics analysis to be 100% reproducible requires a few key elements.
These are:

@@

@@

## Source

Source: https://github.com/markziemann/udocker_r_example

Alt titles:

* "Recipes for extremely reproducible enrichment analysis"

>>>>

---
title: "A recipe for extremely reproducible enrichment analysis"
author: "Mark Ziemann and Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

@@

The lack of reproducibility is particularly acute for enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script based methods
[5](Wijesooriya et al, 2022).
Web-based tools are a potential concern, as it is known that algorithms and functional annotation
sets undergo regular updates, but the historical versions are typically not archived, nor
are versions regularly reported in publications, making later replication difficult if not
impossible in most cases [5](Wijesooriya et al, 2022).
Indeed, in some cases the entire web resource can become unavailable, and hence irreproducible due to
a phenomenon known as link decay [8](Hennessey & Ge 2013).

A lack of reproducibility does not mean a study is invalid, however a lack of auditability means
it cannot be ruled out that mistakes have been made in the enrichment analysis and that the results
are biased or false.
As omics data are becoming more routine in the drug development process and are used to
inform clinical practice, reproducibility should become a minimum requirement for
bioinformatics analysis of omics data.

If research data and code are shared, reproducibility still cannot be guaranteed due to the
differences in various computing environments such as Windows, Linux, Mac, which provide varying
sets of libraries for programming languages to use.
Even if the operating system is specified in the research methods, reproducibility can be difficult if the
operating system is old or obscure, and where dependancies are unavailable.
To address this, many articles have been written to endorse the use of "virtual machines" or "Docker containers"
to encapsulate the operating system and dependancies along with the research code to facilitate
a high degree of reproducibility [9-20](Stein 2010; Dudley & Butte 2010; Sandve et al, 2013;
Piccolo & Frampton 2016; Grüning et al, 2014; Lewis et al, 2016; da Veiga Leprevost et al, 2017;
Kulkarni et al, 2018; Pasquier et al, 2018; Hung et al, 2019; Brito et al, 2020; Piccolo et al, 2021).

Despite the availability of such tools for over a decade and the outsized impact such analysis has
on the conclusions of an omics study, enrichment analyses using fully reproducible virtual machines
or containers remains a miniscule fraction of the overall share of enrichment analyses.
A small number of enrichment analysis applications have been developed to have beneficial properties
of being "containerised" and with archived hostorical versions [21,22](Ge et al, 2020; Perampalam & Dick 2020).
This helps reproducibility greatly, but is insufficient as it lacks information about how the input data was
used and a record of any parameter selections.

Therefore the most complete reproducibility solution is to use containerisation to provide a scripted
workflow with prescribed links to research data, external databases and complete record of parameter selection.
Designed correctly, these Docker images could encapsulate bioinformatics workflows and ensure their
computational reproducibility for decades to come, with just one or a few simple commands.

The barriers to entry include:

* The fact that conventional web based and GUI tools are easier to use

* It adds an overhead of effort to the project

* There's no time to learn new stuff

* It's not a requirement of the journal

We address this by:

* providing working templates for docker image and Rmarkdown scripts for common transcriptome and enrichment analysis
routines.

* Providing step-by-step guides to help users customise these templates for their own work, deploy the analysis,
and verify the results.

To highlight the need for extremely reproducible enrichment analysis, we first investigate the computational
reproducibility of enrichment analyses shown in a sample of 20 journal articles published in 2019.

## Methods

### A systematic assessment of reproducibility in enrichment analysis

<Anusuiya to populate>

### A container based approach to addressing reproducibility of enrichment analysis

Here describe the set up of the system.
The following text is a bit random and needs to be culled.
Although it doesn't need to be this way due to the precise nature of modern computers and software.
In order for bioinformatics analysis to be 100% reproducible requires a few key elements.
These are:

@@

## Concluding remarks

@@

## Bibliography

1.  Slonim DK. From patterns to pathways: gene expression data analysis comes of age. Nat Genet [Internet]. 2002 [cited 2023 Mar 3];32 Suppl(S4):502–8. Available from: https://www.nature.com/articles/ng1033z
2.  Khatri P, Sirota M, Butte AJ. Ten years of pathway analysis: current approaches and outstanding challenges. PLoS Comput Biol [Internet]. 2012 [cited 2023 Mar 3];8(2):e1002375. Available from: https://pubmed.ncbi.nlm.nih.gov/22383865/
3.  Timmons JA, Szkop KJ, Gallagher IJ. Multiple sources of bias confound functional enrichment analysis of global -omics data. Genome Biol [Internet]. 2015;16(1):186. Available from: http://dx.doi.org/10.1186/s13059-015-0761-7
4.  Reimand J, Isserlin R, Voisin V, Kucera M, Tannus-Lopes C, Rostamianfar A, et al. Pathway enrichment analysis and visualization of omics data using g:Profiler, GSEA, Cytoscape and EnrichmentMap. Nat Protoc [Internet]. 2019;14(2):482–517. Available from: http://dx.doi.org/10.1038/s41596-018-0103-9
5.  Wijesooriya K, Jadaan SA, Perera KL, Kaur T, Ziemann M. Urgent need for consistent standards in functional enrichment analysis. PLoS Comput Biol [Internet]. 2022;18(3):e1009935. Available from: http://dx.doi.org/10.1371/journal.pcbi.1009935
6.  Zhao K, Rhee SY. Interpreting omics data with pathway enrichment analysis. Trends Genet [Internet]. 2023 [cited 2023 Mar 1];0(0). Available from: https://www.cell.com/trends/genetics/fulltext/S0168-9525(23)00018-5?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS0168952523000185%3Fshowall%3Dtrue
7.  Peng RD. Reproducible research in computational science. Science [Internet]. 2011;334(6060):1226–7. Available from: http://dx.doi.org/10.1126/science.1213847
8.  Hennessey J, Ge S. A cross disciplinary study of link decay and the effectiveness of mitigation techniques. BMC Bioinformatics [Internet]. 2013;14 Suppl 14(S14):S5. Available from: http://dx.doi.org/10.1186/1471-2105-14-S14-S5
9.  Stein LD. The case for cloud computing in genome informatics. Genome Biol [Internet]. 2010;11(5):207. Available from: http://dx.doi.org/10.1186/gb-2010-11-5-207
10.  Dudley JT, Butte AJ. In silico research in the era of cloud computing. Nat Biotechnol [Internet]. 2010;28(11):1181–5. Available from: http://dx.doi.org/10.1038/nbt1110-1181
11.  Sandve GK, Nekrutenko A, Taylor J, Hovig E. Ten simple rules for reproducible computational research. PLoS Comput Biol [Internet]. 2013;9(10):e1003285. Available from: http://dx.doi.org/10.1371/journal.pcbi.1003285
12.  Piccolo SR, Frampton MB. Tools and techniques for computational reproducibility. Gigascience [Internet]. 2016;5(1). Available from: http://dx.doi.org/10.1186/s13742-016-0135-4
13.  Grüning B, Chilton J, Köster J, Dale R, Soranzo N, van den Beek M, et al. Practical computational reproducibility in the life sciences. Cell Syst [Internet]. 2018;6(6):631–5. Available from: http://dx.doi.org/10.1016/j.cels.2018.03.014
14.  Lewis J, Breeze CE, Charlesworth J, Maclaren OJ, Cooper J. Where next for the reproducibility agenda in computational biology? BMC Syst Biol [Internet]. 2016;10(1). Available from: http://dx.doi.org/10.1186/s12918-016-0288-x
15.  da Veiga Leprevost F, Grüning BA, Alves Aflitos S, Röst HL, Uszkoreit J, Barsnes H, et al. BioContainers: an open-source and community-driven framework for software standardization. Bioinformatics [Internet]. 2017;33(16):2580–2. Available from: http://dx.doi.org/10.1093/bioinformatics/btx192
16.  Pasquier T, Lau MK, Han X, Fong E, Lerner BS, Boose ER, et al. Sharing and Preserving Computational Analyses for Posterity with encapsulator. Comput Sci Eng [Internet]. 2018;20(4):111–24. Available from: https://ieeexplore.ieee.org/abstract/document/8409369
17.  Kulkarni N, Alessandrì L, Panero R, Arigoni M, Olivero M, Ferrero G, et al. Reproducible bioinformatics project: a community for reproducible bioinformatics analysis pipelines. BMC Bioinformatics [Internet]. 2018;19(Suppl 10):349. Available from: http://dx.doi.org/10.1186/s12859-018-2296-x
18.  Hung LH, Hu J, Meiss T, Ingersoll A, Lloyd W, Kristiyanto D, et al. Building containerized workflows using the BioDepot-workflow-builder. Cell Syst [Internet]. 2019;9(5):508-514.e3. Available from: http://dx.doi.org/10.1016/j.cels.2019.08.007
19.  Brito JJ, Li J, Moore JH, Greene CS, Nogoy NA, Garmire LX, et al. Recommendations to enhance rigor and reproducibility in biomedical research. Gigascience [Internet]. 2020 [cited 2023 Mar 9];9(6). Available from: https://pubmed.ncbi.nlm.nih.gov/32479592/
20.  Piccolo SR, Ence ZE, Anderson EC, Chang JT, Bild AH. Simplifying the development of portable, scalable, and reproducible workflows. Elife [Internet]. 2021;10. Available from: http://dx.doi.org/10.7554/eLife.71069
21.  Ge SX, Jung D, Yao R. ShinyGO: a graphical gene-set enrichment tool for animals and plants. Bioinformatics [Internet]. 2020 [cited 2023 Mar 9];36(8):2628–9. Available from: https://pubmed.ncbi.nlm.nih.gov/31882993/
22.  Perampalam P, Dick FA. BEAVR: a browser-based tool for the exploration and visualization of RNA-seq data. BMC Bioinformatics [Internet]. 2020 [cited 2023 Mar 9];21(1):221. Available from: https://pubmed.ncbi.nlm.nih.gov/32471392/
23.  Lund K, Cole JJ, VanderKraats ND, McBryan T, Pchelintsev NA, Clark W, et al. DNMT inhibitors reverse a specific signature of aberrant promoter DNA methylation and associated gene silencing in AML. Genome Biol [Internet]. 2014;15(8):406. Available from: http://dx.doi.org/10.1186/s13059-014-0406-2
24.  Ziemann M, Kaspi A, El-Osta A. Digital expression explorer 2: a repository of uniformly processed RNA sequencing data. Gigascience [Internet]. 2019;8(4). Available from: http://dx.doi.org/10.1093/gigascience/giz022
